\input{preamble}

\title{Introduction to C}
\author{Eivind Fonn}
\institute{SINTEF ICT / NTNU}
\date{December 2015}
\maketitle

\begin{frame}
  \frametitle{History of C}
  \begin{itemize}
  \item Originated in 1972. By 1973, UNIX was written in C.
  \item K\&R book written in 1978. The first informal specification.
  \item In 1983 ANSI standardized the language, forming what is known as ANSI C.
    It was transferred to ISO in 1990, whence the name C89 or C90. These terms
    all refer to the same language.
  \item The standard was revised in 1999, giving the C99 standard, which is the
    best compromise today between modernity and portability. It is also the
    common subset between C and C++.
  \item C11 is even newer but slightly less portable.
  \item If necessary, specify the standard when compiling: \texttt{-std=c99}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Strengths and weaknesses}
  \begin{itemize}
  \item C is a low-level programming language. Probably the lowest level
    language most programmers will ever use.
  \item This makes it fast and powerful, but not expressive, and not often not
    very convenient.
  \item More so than other languages, C will do what you tell it to, and if you
    screw it up it's your own fault. This gives the programmer a lot of power
    not easily available in other languages.
  \item C has \emph{no automatic memory management}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Strengths and weaknesses}
  \begin{itemize}
  \item C is extremely portable. There are C compilers available for almost
    every architecture.
  \item C's calling convention is standardized. This makes it very easy for
    other languages to call C libraries, and this makes C code even more
    portable.
  \item Almost every modern language ships with an FFI
    (\emph{Foreign Function Interface}) that can call C functions. This gives
    them immediate access to popular libraries (BLAS\ldots).
  \item Most dynamic programming languages (Python, Ruby, Lua\ldots) are written
    in C and allow extension modules written in C (e.g. Numpy).
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Variables}
  \begin{itemize}
  \item Unlike e.g. Python, which has \emph{names}, C has genuine variables.
    That is, each variable is associated with a name and a \emph{location}.
  \item Declare a variable by giving it a type and a name:
\begin{lstlisting}[style=c]
  int my_number, other_number;
  double my_real_number;
  char some_byte;
\end{lstlisting}
  \item Variables are not initialized. They have whatever value happen to be in
    the memory location when they are declared. \textbf{Do not use uninitialized
      data.}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Initialize and declare}
  \begin{itemize}
  \item You can initialize and declare a variable at the same time. It's good to
    get into the habit of doing this so you never accidentally use an
    uninitialized variable.
\begin{lstlisting}[style=c]
  int my_number = 42;
  double my_real_number = 3.14;
  char some_byte = long_expression();
\end{lstlisting}
  \item Compile your code with \texttt{-Wuninitialized} to get warnings when you
    appear to be using uninitialized variables.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Types}
  The most common types are
  \begin{itemize}
  \item \texttt{int}: the basic signed integer type. The size is
    machine-dependent but is guaranteed to be at least 16 bits. On most
    computers today it will be 32 bits.
  \item \texttt{float}: typically a single-precision floating-point number.
  \item \texttt{double}: typically a double-precision floating-point number.
  \item \texttt{char}: the smallest addressable unit. This will almost always be
    a byte. \textbf{Note:} a \texttt{char} is an \emph{integer} type.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Types}
  \begin{itemize}
  \item Ints and chars optionally can be \emph{signed} or \emph{unsigned}. A
    signed int can (for example) take values in $[-2^{31}, 2^{31})$ while an
    unsigned one can take values in $[0,2^{32})$.
  \item The following code will, on most computers, print ``$200, -56$''.
\begin{lstlisting}[style=c]
  unsigned char a = 200;
  signed char b = 200;
  printf("%d, %d\n", a, b);
\end{lstlisting}
  \item Signed is implied if not otherwise specified.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Types}
  \begin{itemize}
  \item Ints can additionally be either short, normal, long or long long.
  \item Their lengths are guaranteed to be non-decreasing. Shorts and normal
    ints are at least 16 bits, long ints are at least 32 bits and long long ints
    are at least 64 bits.
  \item These qualifiers can be mixed with signed or unsigned. The typename
    \texttt{int} can also be omitted. These four variables have identical types:
\begin{lstlisting}[style=c]
  long a;
  signed long b;
  long int c;
  signed long int d;
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Other types}
  \begin{itemize}
  \item \texttt{size\_t} is an unsigned integer type used for \emph{sizes} of
    things in memory. For example, \texttt{malloc} expects a \texttt{size\_t}
    parameter.
  \item \texttt{ptrdiff\_t} is a signed integer type used for differences
    between memory locations.
  \item Both of these are useful for indexing arrays. Where only nonnegative
    indices are expected (which is almost always the case), I prefer
    \texttt{size\_t}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Other types}
  Other types with predictable and portable sizes were introduced in C99: Here,
  \texttt{N} can be $8$, $16$, $32$ or $64$.
  \begin{itemize}
  \item \texttt{[u]intN\_t}: integer type with exact width.
  \item \texttt{[u]int\_leastN\_t}: integer type with at least a given width.
  \item \texttt{[u]int\_fastN\_t}: the \emph{fastest} type on the given machine
    with at least a given width.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Arithmetic}
  \begin{itemize}
  \item For both signed and unsigned integers, arithmetic in C may overflow
    without warning if the result is too big or too small.
  \item Mathematics works modulo $2^n$ where $n$ is the size of the type in
    question.
  \item The usual rules of precedence apply to \texttt{+}, \texttt{-},
    \texttt{*} and \texttt{/}. Associativity is to the left. Therefore,
\begin{lstlisting}[style=c]
  arr[i] = 1.0 / ((i+1) * (i+1));
\end{lstlisting}
    is better written as
\begin{lstlisting}[style=c]
  arr[i] = 1.0 / (i+1) / (i+1);
\end{lstlisting}
    This also avoids overflow!
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Arrays}
  \begin{itemize}
  \item Arrays can be declared and initialized
\begin{lstlisting}[style=c]
  int cat[4];  // Uninitialized
  int cat[4] = {1, 2, 3, 4};
  int cat[] = {1, 2, 3, 4};
  int cat[];  // Error
\end{lstlisting}
  \item Since C99, the array size may be known only at runtime. This is an error
    in ANSI C:
\begin{lstlisting}[style=c]
  int size = 4;
  int cat[size];  // Uninitialized
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Arrays}
  \begin{itemize}
  \item Multidimensional arrays are stored contiguously in memory, in row-major
    order.
  \item For this reason, you can only leave the first dimension unspecified.
    Otherwise the compiler doesn't know how to compute the memory location of an
    array element from its index.
\begin{lstlisting}[style=c]
  int cat[4][4];
  int cat[4][4] = { { ... }, { ... }, ... };
  int cat[][4] = { { ... }, { ... }, ... };
  int cat[][] = ...;  // Error
  int cat[][4];  // Error
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Pointers}
  \begin{itemize}
  \item Pointers are memory addresses. Encoded in the type of a pointer is also
    the type of the thing it points to.
\begin{lstlisting}[style=c]
  int *a;  // Pointer to int
  double *b;  // Pointer to double
  float **c;  // Pointer to pointer to float
\end{lstlisting}
  \item To get the value that a pointer points to, use dereferencing with
    \texttt{*}. To get the address of a variable, use \texttt{\&}:
\begin{lstlisting}[style=c]
  int a = 10;
  int *b = &a;
  printf("b points to the value %d\n", *b);
  printf("a has address %x\n", &a);
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Pointers}
  \begin{itemize}
  \item It is easy to confuse the \texttt{*} in the declaration and the
    dereferencing operator.
  \item Helpful way of thinking: \texttt{int *a;} declares \texttt{*a} as an
    \texttt{int}. This is why C people like to put the \texttt{*} next to the
    variable name instead of the type.
  \item \textbf{Note:} this is just a mnemonic device. It does not actually
    declare an int.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Pointers}
  \begin{itemize}
  \item The warning about not using uninitialized memory is extremely important
    for pointers.
  \item There is no guarantee that an uninitialized pointer points to
    somewhere that makes sense, or is even a legal memory location.
  \item Dereferencing such a pointer may cause your program to crash
    (segmentation violation).
  \item It is common to use \texttt{0} or \texttt{NULL} as a value to indicate
    an illegal pointer.
  \item Functions that return pointers may use \texttt{NULL} to indicate an
    error.
  \item Functions that accept pointer arguments may interpret \texttt{NULL} as
    the absence of a value.
  \item \textbf{Be careful!} The only correct answer is also a bad one: check
    every pointer everywhere.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Pointers as arrays}
  \begin{itemize}
  \item Pointers are unsigned integers and can be subject to arithmetic.
  \item The math is always in terms of the underlying type.
\begin{lstlisting}[style=c]
  int a = 10;
  int *b = &a + 1;
\end{lstlisting}
    The pointer \texttt{b} now points to a memory location \emph{four bytes}
    offset from where \texttt{a} is stored (if ints are four bytes).
  \item Therefore we can use pointers as a form of arrays.
\begin{lstlisting}[style=c]
  int *ptr = ...;
  *ptr;        // First element
  *(ptr + 1);  // Second element
  *(ptr + 2);  // Third element
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Pointers as arrays}
  \begin{itemize}
  \item This is so useful that C allows us to use \texttt{[]}-indexing on
    pointers.
  \item If \texttt{p} has a pointer type then \texttt{p[n]} is
    \emph{exactly equivalent} to \texttt{*(p+n)}.
  \item Interestingly, since \texttt{p+n} is the same as \texttt{n+p}, you can
    also use \texttt{n[p]}.
  \item This has caused a lot of confusion about the difference between pointers
    and arrays.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Pointers as arrays (1D)}
\begin{lstlisting}[style=c]
  int *ptr;
  int arr[];
  ptr[5]; arr[5];
\end{lstlisting}
  \begin{itemize}
  \item For \emph{one-dimensional} arrays, array indexing and pointer indexing
    work pretty much exactly the same way.
  \item Array indexing may be slightly faster since it involves less redirection
    (For arrays, the program gets ``for free'' the location of the first
    element. In the pointer case, the location of the first element is something
    that's stored in a variable which has to be looked up.)
  \item When passed to a function, \textbf{arrays become pointers}.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Pointers as arrays (1D)}
\begin{lstlisting}[style=c]
  int **ptr;
  int arr[][8];
  ptr[5][3]; arr[5][3];
\end{lstlisting}
  \begin{itemize}
  \item For \emph{multi-dimensional} arrays, they are very different things.
  \item Array: get the element at offset $5\times8 + 3 = 43$ from the start.
  \item Pointer: Follow the pointer, go to offset $5$, follow \emph{that}
    pointer, go to offset $3$.
  \item When passing a multi-dimensional array to a function, only the ``top''
    layer becomes a pointer. The other layers of redirection can't be
    automatically created.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Booleans}
  \begin{itemize}
  \item C notably does not have a boolean type.
  \item Instead it uses integers. Anything that is not zero is true.
  \item Pitfall:
\begin{lstlisting}[style=c]
  unsigned char b = 256;
  if (b) { ... };
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Strings}
  \begin{itemize}
  \item Also notably, C does not have a string type.
  \item Instead, we use pointers to chars.
  \item A C string implicitly keeps going until a zero byte is found. This byte
    is not part of the string but it is part of the allocated memory for it.
  \item The meaning of the \texttt{char **argv} argument to \texttt{main} is
    then to be understood as an ``array'' of ``strings''.
  \item \texttt{strcmp} compares two strings. It returns zero if they are equal.
    Therefore it should be negated in an \texttt{if}.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Structs}
  \begin{itemize}
  \item A struct is a handy way of grouping variables together.
\begin{lstlisting}[style=c]
  struct person {
    int age;
    char *name, *address;
    double height, weight;
  };

  struct person me;
  me.name = "Eivind Fonn";
  me.age = 31;
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Structs}
  \begin{itemize}
  \item We often find ourselves with pointers to structs.
  \item This is ugly:
\begin{lstlisting}[style=c]
  struct person *me;
  (*me).age;
\end{lstlisting}
  \item Instead, use \texttt{->} as a shortcut.
\begin{lstlisting}[style=c]
  struct person *me;
  me->age;
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Scope}
  \begin{itemize}
  \item Each variable has a \emph{scope} in which it lives.
  \item In C, a variable is scoped to the closest surrounding block of braces:
    \texttt{\{} and \texttt{\}}. When that block ends, the variable dies.
  \item Pointers are destroyed but the pointed-to values are not!
  \item C is lexically scoped: a symbol may refer to a variable in this scope or
    any parent scope. Parent scopes are determined according to code
    \emph{as written} and not as functions are \emph{called}. That means a
    called function can't access variables in the scope of the caller. (Yes,
    such languages exist.)
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Scope and MPI}
  This has interesting consequences when writing MPI code. The following will
  not work:
\begin{lstlisting}[style=c]
  if (rank == 0) {
    // This variable is only needed on proc 0
    int something;
  }

  // Code that needs to run everywhere

  if (rank == 0) {
    // Error!
    something = ...;
  }
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Functions}
  \begin{itemize}
  \item Look very similar to functions in other C-like languages.
\begin{lstlisting}[style=c]
  <rtype> <name>(<arguments>)
  {
    <code>
    return <expr>;
  }
\end{lstlisting}
  \item Arguments are always passed by value. If you want to pass by reference,
    to modify the arguments in a way that is visible to the caller, you must use
    pointers.
  \item This is a handy way to return more than one value.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Functions}
  For example:
\begin{lstlisting}[style=c]
  int sum(int a, int b) {
    return a + b;
  }
  int product(int a, int b) {
    return a * b;
  }
  void sum_and_product(int a, int b,
                       int *sum, int *product)
  {
    *sum = a + b;
    *product = a * b;
  }
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Functions}
  \begin{itemize}
  \item A function must be \emph{declared} before it is \emph{used} in a source
    code file. The declaration includes all information the compiler needs to
    use the function:
\begin{lstlisting}[style=c]
    int sum(int a, int b);
    int sum(int, int);
\end{lstlisting}
  \item As long as you have provided a \emph{definition} when the program is
    linked, it's okay.
  \item This lets us split up code into different source code files, so long as
    we keep the declarations separate (in header files).
  \item When you \texttt{\#include} files, you're importing a bunch of function
    declarations.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Arrays, pointers and functions}
\begin{lstlisting}[style=c]
  void fncA(int *a);
  void fncB(int b[]);  // Equivalent to fncA

  void fncC(int **c);
  void fncD(int d[][8]);  // NOT equivalent to fncC
  void fncE(int (*e)[8]);  // Equivalent to fncD

  int arr_1d[10];
  fncA(arr_1d); fncB(arr_1d);  // OK

  int arr_2d[10][8];
  fncC(arr_2d);  // Not OK
  fncD(arr_2d); fncE(arr_2d);  // OK
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Function pointers}
  \begin{itemize}
  \item You can store a pointer to a function.
  \item This declares \texttt{func\_ptr} as a pointer to a function that accepts
    a char and a pointer to a double, and returns an int. Then it sets this
    pointer to the address of the function \texttt{actual\_func}.
\begin{lstlisting}[style=c]
  int (*func_ptr)(char, double*) = &actual_func;
\end{lstlisting}
  \item You can call a function pointer by dereferencing it.
\begin{lstlisting}[style=c]
  int value = (*func_ptr)('a', &some_double);
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Typedefs}
  \begin{itemize}
  \item C types can be famously complicated to figure out.
\begin{lstlisting}[style=c]
  void (*f)(int,void (*)());
  int **(*f)(int**,int**(*)(int**,int**));
  char (*(*X[3])())[5];
\end{lstlisting}
  \item Helpful tool: cdecl
\begin{lstlisting}
  cdecl> explain void (*f)(int,void(*)())
  declare f as pointer to function
  (int, pointer to function returning void)
  returning void
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Typedefs}
  \begin{itemize}
  \item Better technique: declare type aliases with \texttt{typedef}.
  \item Very useful for structs so you don't have to type \texttt{struct person}
    every time.
\begin{lstlisting}[style=c]
  typedef struct person person_t;
\end{lstlisting}
  \item You can typedef together with the declaration too.
\begin{lstlisting}[style=c]
  typedef struct {
    int age;
    char *name, *address;
    double height, weight;
  } person_t;
\end{lstlisting}
  \item Also useful for types that will always be pointers.
\begin{lstlisting}[style=c]
  typedef person_t* personPtr;
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Constants}
  \begin{itemize}
  \item A variable can be declared \texttt{const} to indicate that its value
    will not change.
  \item In function arguments, \texttt{const} can be used with pointers to
    indicate that the function will not change the data it points to.
  \item The latter allows compiler optimization, so use it when possible!
  \item Easy to get confused, though:
\begin{lstlisting}[style=c]
  int const *ptr;  // *ptr is constant
  int * const ptr;  // ptr is constant
  int const * const ptr;  // both
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Preprocessor directives}
  \begin{itemize}
  \item You can use \texttt{\#define} to define symbolic constants. These are
    substituted textually before compilation, so they're effectively the same as
    writing a literal value in your program:
\begin{lstlisting}[style=c]
  #define PI 3.141516;
  return r * PI * PI;

  // turns into
  return r * 3.141516 * 3.141516;
\end{lstlisting}
  \item Note that they are not variables, not even const variables.
\begin{lstlisting}[style=c]
  &PI;  // Error
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Preprocessor directives}
  \begin{itemize}
  \item You can use \texttt{\#define} in a more general sense for macros:
\begin{lstlisting}[style=c]
  #define max(a,b) a > b ? a : b
  return max(x, 0);

  // turns into
  return x > 0 ? x : 0;
\end{lstlisting}
  \item Beware:
\begin{lstlisting}[style=c]
  #define prod(a,b) a * b
  return prod(x + y, r + s)

  // turns into
  return x + y * r + s;
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Preprocessor directives}
  \begin{itemize}
  \item Therefore, when defining macros, there is no such thing as too many
    parentheses.
\begin{lstlisting}[style=c]
  #define max(a,b) ((a) > (b) ? (a) : (b))
  #define prod(a,b) ((a) * (b))
\end{lstlisting}
  \item Just as \texttt{\#define} ``constants'' are not real constants, macros
    are \emph{not} functions.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Preprocessor directives}
  \begin{itemize}
  \item You can use preprocessor directives to selectively enable or disable
    features in a program.
\begin{lstlisting}[style=c, basicstyle=\ttfamily\scriptsize]
  #ifdef HAVE_MPI
  #ifdef HAVE_OPENMP
    int acquired;
    MPI_Init_thread(&argc, &argv,
                    MPI_THREAD_FUNNELED,
                    &acquired);
    MPI_Comm_rank(MPI_COMM_WORLD, rank);
    if (*rank == 0)
      printf("Acquired MPI threading level: %d\n", acquired);
  #else
    MPI_Init(&argc, &argv);
  #endif
    MPI_Comm_size(MPI_COMM_WORLD, size);
    MPI_Comm_rank(MPI_COMM_WORLD, rank);
  #else
    *rank = 0;
    *size = 1;
  #endif
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Preprocessor directives}
  \begin{itemize}
  \item Compilers allow you to define preprocessor symbols when compiling.
    E.g. \texttt{gcc -DHAVE\_MPI ...}, or with CMake:
\begin{lstlisting}
  add_definitions(-DHAVE_MPI)
\end{lstlisting}
  \item A common trick for making sure a header file is only included once:
\begin{lstlisting}
  #ifndef _MY_HEADER_FILE_H_
  #define _MY_HEADER_FILE_H_

  // ...

  #endif
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Miscellaneous tips}
  \begin{itemize}
  \item \texttt{malloc} and \texttt{calloc} allocate memory and return pointers.
    The latter zeroes the allocated memory, which makes it safer but also
    slower.
  \item \texttt{sizeof} is a compile-time macro for getting the size of a type.
    It is useful together with \texttt{malloc} and \texttt{calloc}.
  \item If you allocate memory but ``lose'' the pointer to it, that memory
    cannot be used by anything else until your program exits. Use \texttt{free}
    to release memory you don't need any more.
  \item Use \texttt{atoi} and \texttt{atof} to convert strings (e.g.
    command-line arguments) to integers or doubles.
  \item Use \texttt{printf} to convert to string (and output).
  \item Use casting for other automatic conversions: \texttt{(double)some\_int}.
  \item Just because it's easy to mess up in C doesn't mean you should litter
    your program with safety checks. We're not NASA.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Miscellaneous tips}
  \begin{itemize}
  \item Store multidimensional arrays contiguously. This lets us use BLAS. If
    you \texttt{malloc} inside a loop you may be doing it wrong. You can
    construct an array of pointers afterwards, if you like.
\begin{lstlisting}[style=c]
  double *data =
    malloc(M * N * sizeof(double));
  double *matrix[N];
  for (size_t i = 0; i < N; i++) {
    matrix[i] = &data[i*M];
    // or
    matrix[i] = data + i*M;
  }
\end{lstlisting}
  \end{itemize}
\end{frame}

\input{postamble}
